Implementation of a lattice Boltzmann method for immiscible two-phase flow simulations using the level set method

### 2015-11-15 Paper/paper.tex
Referenz fuer NSE

In this publication we use a hybrid lattice Boltzmann level set method. The motion of the interface is modelled by the level set method. The input data for the level set method is given by the lattice Boltzmann method (LBM), which is applied independently for each fluid. This method has the advantage that it allows large ranges of viscosity- and density-ratios between the fluid phases, while describing a sharp inteface.

The incompressible two-phase flow problem can be described mathematically with the Navier-Stokes equations for two incompressible fluids (\cref{nse1,nse2}). 
%TODO Regina: Hier fehlen Referenzen!
To apply them, we distribute our domain into three parts: $\Omega_1$ and $\Omega_2$ describe the two fluids and $\Gamma = \delta\Omega_1 \cap \delta\Omega_2$ is the interface inbetween (\cref{intro_phases}). The material properties in each fluid are constant and the interface is assumed to be sharp. Thus, the overall material properties are discontinuous and the fluids will not mix.

@@

Let $\Gamma_t$ be our interface at time t. Within the level set method, $\Gamma_t$ is the zero iso-surface of the level set function $\varphi$. Thus, for $\vec x \in \Gamma_t$ we have $\varphi(\vec x,t) = 0$, in direction of the surface normal we get $\varphi > 0$, and in the other direction we get $\varphi < 0$. For this level set function we get the level set equation introduced by Osher and Sethian in \cite{OsherSethian} for the evolution of surfaces,
\begin{equation}
  \frac{\partial \varphi}{\partial t} + \vec v \cdot \nabla \varphi = 0 \text{,}
\end{equation}
with the velocity $\vec v = \vec u(\vec x(t))$ for $\vec x(t) \in \Gamma_t$, which we obtain from the LBM.

@@

With respect to the given definition of the discrete lattice velocities (\ref{disc_vel}, see also \cite{Thoemmes2}), the pressure jump is approximated by,
\begin{equation}
	[p] \approx \frac{1}{3} (\rho(x_2,t)\varrho_2 - \rho(x_1,t)\varrho_1) \text{.}
\end{equation}
and thus, all variables for the boundary conditions are defined. % sagen wir hier noch, dass im Thömmes-paper ein Fehler war, der im zweiten Paper korrigiert wurde?

@@

Figure \ref{couette2} shows a second scenario with $5 \times 10$ cells along $x \times y$ direction and viscosities $\mu^{(1)} = 1$ and $\mu^{(2)} = 0.2$ with the interface located at $y=0.75$. Again the numerically approximated values agree very well with the analytic ones and the simulation converged at a relative error of $1.4\cdot10^{-4}$ after 2000 LBM iterations.
\begin{figure}[t!]
	\flushright
	\hfill\includegraphics[trim = 0mm 8cm 0mm 8cm, clip, width=.8\textwidth, natwidth=595,natheight=842]{couette2.pdf}\hspace*{\fill}
	\caption{Two fluid domains $\Omega_i$ separated by an interface.}
	\label{couette2}
\end{figure}
We conclude from this that the implemented LBM boundary conditions successfully model viscosity jumps at the fluid interface.
\subsection{Young-Laplace experiment for a bubble}
Our Couette flow scenario does not account for the influence of surface tension as the fluid interface is not curved. If it would get distorded, surface tension would have a stabilizing effect although. 
Hence, we use the Young-Laplace law for the pressure drop across curved surfaces, to validate the implementation of surface tension. The pressure inside of a convex surfaces is higher than on the outside and the pressure drop $\Delta p$ of a bubble of radius $r$ is proportional to the surface tension coefficient $\sigma$ according to thefollowing relationship,
\begin{equation}
	\Delta p = \frac{2\sigma}{r} = 2 \sigma \kappa .
\end{equation}

@@

@@

\bibitem{LBM4} {\sc D.A. WOLF-GLADOW}, Lattice-Gas Cellular Automata and Lattice Boltzmann Models, An Introduction, Springer, 2000.

>>>>

In this publication we use a hybrid lattice Boltzmann level set method. The motion of the interface is modelled by the level set method. The input data for the level set method is given by the lattice Boltzmann method (LBM), which is applied independently for each fluid. This method has the advantage that it allows large ranges of viscosity- and density-ratios between the fluid phases, while describing a sharp interface.

The incompressible two-phase flow problem can be described mathematically with the Navier-Stokes equations for two incompressible fluids (\cref{nse1,nse2}, \cite{ferziger}). 
%TODO Regina: Hier fehlen Referenzen!
To apply them, we distribute our domain into three parts: $\Omega_1$ and $\Omega_2$ describe the two fluids and $\Gamma = \delta\Omega_1 \cap \delta\Omega_2$ is the interface in-between (\cref{intro_phases}). The material properties in each fluid are constant and the interface is assumed to be sharp. Thus, the overall material properties are discontinuous and the fluids will not mix.

@@

Let $\Gamma_t$ be our interface at time t. Within the level set method, $\Gamma_t$ is the zero isosurface of the level set function $\varphi$. Thus, for $\vec x \in \Gamma_t$ we have $\varphi(\vec x,t) = 0$, in direction of the surface normal we get $\varphi > 0$, and in the other direction we get $\varphi < 0$. For this level set function we get the level set equation introduced by Osher and Sethian in \cite{OsherSethian} for the evolution of surfaces,
\begin{equation}
  \frac{\partial \varphi}{\partial t} + \vec v \cdot \nabla \varphi = 0 \text{,}
\end{equation}
with the velocity $\vec v = \vec u(\vec x(t))$ for $\vec x(t) \in \Gamma_t$, which we obtain from the LBM.

@@

With respect to the given definition of the discrete lattice velocities (\ref{disc_vel}, see also \cite{Thoemmes2}), the pressure jump is approximated by,
\begin{equation}
	[p] \approx \frac{1}{3} (\rho(x_2,t)\varrho_2 - \rho(x_1,t)\varrho_1) \text{.}
\end{equation}
and thus, all variables for the boundary conditions are defined.

@@

Figure \ref{couette2} shows a second scenario with $5 \times 10$ cells along $x \times y$ direction and viscosities $\mu^{(1)} = 1$ and $\mu^{(2)} = 0.2$ with the interface located at $y=0.75$. Again the numerically approximated values agree very well with the analytic ones and the simulation converged at a relative error of $1.4\cdot10^{-4}$ after 2000 LBM iterations.
\begin{figure}[t!]
	\flushright
	\hfill\includegraphics[trim = 0mm 8cm 0mm 8cm, clip, width=.8\textwidth, natwidth=595,natheight=842]{couette2.pdf}\hspace*{\fill}
	\caption{Two fluid domains $\Omega_i$ separated by an interface.}
	\label{couette2}
\end{figure}
We conclude from this that the implemented LBM boundary conditions successfully model viscosity jumps at the fluid interface.
\subsection{Young-Laplace experiment for a bubble}
Our Couette flow scenario does not account for the influence of surface tension as the fluid interface is not curved. If it would get distorted, surface tension would have a stabilizing effect although. 
Hence, we use the Young-Laplace law for the pressure drop across curved surfaces, to validate the implementation of surface tension. The pressure inside of a convex surfaces is higher than on the outside and the pressure drop $\Delta p$ of a bubble of radius $r$ is proportional to the surface tension coefficient $\sigma$ according to the following relationship,
\begin{equation}
	\Delta p = \frac{2\sigma}{r} = 2 \sigma \kappa .
\end{equation}

@@

\bibitem{ferziger} {\sc J. FERZIGER, M. PERIC.}, Computational methods for fluid dynamics. Springer Science \& Business Media, ISBN 978-3540420743 (2012).

@@

\bibitem{LBM4} {\sc D.A. WOLF-GLADOW}, Lattice-Gas Cellular Automata and Lattice Boltzmann Models, An Introduction, Springer, (2000).
